Optimize sequencing depth for shotgun metagenomics of pollination system by rarefaction, using a modular profiling pipeline

### 2021-09-08 MResProject/Latex/Draft.tex
CMEECourseWork

>>>>

\subsection{Presentation}
As pollinators, bees play crucial role in agriculture and provide huge economic benefit
but their populations are facing challenges from factors like invasive species, parasites, pesticide exposure, habitat loss and climate change. 
Together, these make bee health a crucial issue.
Like other animals, bees are associated with diverse gut microbial communities impacting host health. 
It includes bacteria, fungi and viruses. 
It also provides materials of environmental DNA, indicating interactions with other organisms.
This microbiome influences bee health significantly by mediating processes including food digestion, parasite defence and chemical detoxification. 
Exploration of bee microbiome has been boosted by high throughput methods such as amplicon sequencing and shotgun metagenomics.
Amplicon sequencing captures species-specific region to reveal composition of microbial communities. 
It is cost-effective and is vastly used. 
However, it does not cover all taxa and does not provide information on gene content, making unreliable function inference. (strain diversity)
(Strains of Gilliamella apicola variant in metabolizing mannose, xylose, arabinose, and rhamnose, all of which are toxic to bees)
(Strains of lactobacilli and bifidobacteria are highly variant in carbohydrate metabolism)
Shotgun metagenomics provides a solution to overcome these drawbacks by capturing DNA fragments unselectively.
However, shotgun metagenomics is of complex data analysis and high cost.   
Generally, one of the first steps of data analysis is assembling short reads into long contigs, which enable accurate taxonomic/functional annotation. 
But assembly is of high false negative rate since there are always reads left unassembled. 
Alternatively, short reads can also be processed directly by mapping to a reference database, but the output might be inaccurate due to limited mappability.
Besides, shotgun metagenomics is high cost. 
A shallow 6 Gbp metagenomic dataset may cost 120 pounds per sample, while a standard 15 Mbp amplicon sequencing only costs about 20 pounds per sample. 
Here, I aimed to address data analysis pipeline and high cost shotgun metagenomics. 
I developed an integrated pipeline combining assembly-dependent and assembly-free methods. 
Raw reads are first filtered to remove low quality data and host contamination. 
Then clean reads are assembled for species identification. 
To resolve species not represent by the assembly, I mapped reads to genomes of identified species. 
Unmapped reads are processed directly for species identification. 
Using this pipeline, I processed datasets from hony bees, bumble bees and flower washes.
Although these datasets are different in proportion of host contamination, they all represent a multi-kingdom community.
Bees possess a highly conserved and specialized core gut microbiome, which consists of bacteria including Bifidobacterium, Lactobacillus, Gilliamella, Frischella and Snodgrassella. 
This heatmap shows the composition of bacterial communities and as expected, a complete set of core bacterial were found. 
It indicates these samples were of good quality and representivity.
As for other species, there were insect pollinators and crops. Fungal communities were dominated by Nosema, a widespread bee pathogen and some yeast species. 
Then I compared integrated pipeline with a standard assembly-dependent methods. 
These points above zero means the integrated pipeline resolved species not represented by assembly
A brief summary: I developed a pipeline analyzing metagenomic data. 
It is capable of finding more species than assembly-dependent methods. 
Besides, it is easy to incorporate alternative tools into each module and is easy for troubleshooting since all output files are well documented. 
Then I optimized sequencing depth for metagenomics of different host species, in order to define a trade-off between sequencing effort and reliable species/function diversity analysis.
First, I measured species/function diversity. 
There are several diversity indexes used in ecology and among them, I picked Hill numbers. 
Hill number is a family of parametric diversity indexes and it provides several unique advantages. 
First, it obeys replication principle: 
Assume there are two equally diverse assemblages with no shared species, the diversity of pooled assemblage is twice the single assemblages. 
It indicates that as sequencing depth increase, discovery of novel species would lead to increase in Hill numbers. 
When the sequencing depth is so big that almost all species have been found, Hill numbers would level off. 
Second, its sensitivity to relative abundances can be modulated by values of order. 
Big order value means little emphasis on rare species. 
This is useful for metagenomics because rare species are more likely to be false positives than abundant species.
Finally, it can be transformed to other diversity indexes like richness, shannon index and simpson index through simple algebraic transformation.
So I simulated shallow sequencing by randomly subsampling the original datasets and fitted the relationship between Hill numbers and sequencing depth. 
The slope of the curve reflects extra diversity brought by additional sequencing effort. 
So I looked the point at which the curve slope is small enough and defined it as the optimal sequencing depth, a trade-off between sequencing cost and obtaining meaningful information on species/function diversity. 
This analysis assumes that the original dataset is almost complete, namely almost all species or function category are represented by the original dataset. 
So I first looked at relationship between richness, or Hill number of order zero, and sequencing depth. 
If the original dataset is complete, the model should level off at the final point. 
Figures in first row are for species richness. 
One honey bee sample and the flower sample were dropped because of incompleteness. 
Figures in second row are for function diversity and no curve levels off.
As a result, 
for species diversity estimation, 12.10 Gbp for honey bees and 12.75 Gbp for bumble bees would be sufficient. 
Sequencing depth can be further reduced if there is little emphasis on rare species. 
As for functional diversity, no datasets were complete.
So I estimated optimal sequencing depth for species diversity estimation of honey bee and bumble bee samples. 
For function diversity, even the deepest dataset was not sufficient, indicating more sequencing effort is needed.
Finally some take-home message. 
I developed a pipeline for metagenomic data analysis and it provides improvement in species identification. 
Then I found that about 12 Gbp would be sufficient for species diversity estimation of honey bee and bumble bee samples, and that would be about 200 pounds per sample. 
This sequencing depth can be further reduced if little weight is given to rare species. 
As for function diversity, deep sequencing is valuable. 
Similar experiments can be performed as pilot study of large scale projects and are helpful for budget management.
